---
title: "2. Data types and structures in R"
author: "Alexandr Byzov"
date: "07 09 2020"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# Introduction into objects in R

The most important data structure in R is a vector. Vectors come in two types: atomic vectors and lists.

## Atomic vectors 

There are at least four classes of atomic vectors:

  1. Integers (7L, 5L, 23L);
  2. Double (2.32, 1.19, 4.000);
  3. Logical (TRUE, FALSE, T, F);
  4. Character ('text', '5 in quotation mark is still a character')

To create a vector of each type you can just write it down and press *Ctrl+Enter* on this line or click on *'Run'* in the top right corner of the script window.
```{r}
5
```
Your console have just provided a following output: [1] 5. '[1]' stands for the first element of your output.
```{r}
'text'
```

```{r}
TRUE
```

Missing values in R are represented with *NA* special symbol.
```{r}
NA
```

These vectors are scalar, i.e. they contain only an individual value. To create vectors of two elements or more you need to use a function **c()**.
```{r}
c(1.1, 2, 3.4, 5.55)
```

### Attributes 

Each R object has attributes. They contain metadata on the object. These attribute include:

  1. dimensionality of an object - could be accesed with dim() function
  2. class of an object - could be accesed with typeof() or class()
  3. length of an object - could be accesed with length()
  4. other user-defined attributes/metadata of an object
  
### Implicit coercion 
  
It is crucial to remember that *an atomic vector can only contain objects of the same class*. You can see what happens if you create a vector with mixed classes.

Vector 1:
```{r}
c(TRUE, 1L)
```
Vector 2:
```{r}
c(TRUE, 1L, 2.0)
```
Vector 3:
```{r}
c(TRUE, 1L, 2.0, '3')
```
  
Call function typeof() on each of the aforementioned vectors:
Vector 1:
```{r}
typeof(c(TRUE, 1L))
```
Vector 2:
```{r}
typeof(c(TRUE, 1L, 2.0))
```
Vector 3:
```{r}
typeof(c(TRUE, 1L, 2.0, '3'))
```

As it can be seen, R has coerced data to the most encompassing atomic vector class. This fact will be extremely important throughout this course.

### Explicit coercion

You can also invoke coercion with the following functions: as.integer(), as.double(), as.logical(), as.character()
```{r}
as.logical(c(1, 2, 3, 0))
```

```{r}
as.integer(2.3, 4.5)
```

```{r}
as.character(T, F, 1, 2, 3)
```

### Matrices 

Matrices are an atomic vectors with a dimension attribute. They are computationally efficient, which is why we will use them a lot in this course. Matrices are created with the following function: **matrix()**. This function has several arguments. To understand how each function works, you can invoke its documentation with question mark before function:
```{r}
#?matrix()
```

  1. what does *nrow* argument do?
  2. what does *byrow* argument do?

To create a matrix of 6 elements with two columns and three rows, you need to use at least one additional argument, either nrow or ncol:

```{r}
matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)
```

```{r}
matrix(c(1, 2, 3, 4, 5, 6), ncol = 2)
```

Matrices are filled columnwise by default. To change this behavior, you need to change the value of an argument *byrow*

```{r}
matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, byrow = TRUE)
```

Previously, I have mentioned that matrices have dimensions. To check dimensions of this matrix, you can use function **dim()**
```{r}
dim(matrix(c(1,2,3,4,5,6), nrow = 3, ncol = 2))
```

### [S3 atomic vectors](https://adv-r.hadley.nz/vectors-chap.html#s3-atomic-vectors)

S3 objects are an important topic, however it is a little advanced for now. However, you will rely heavily on one instance of such objects, factors. 

Factor is  a vector with predefined values. In practice, factors are employed to store categorical data. Most of the sociological surveys produce factors.

Factors are built upon integer class. They have an additional attribute - levels. To create a factor, you need to call function *factor()*. Call the documentation for this function and try to specify a sample of 2 men and 4 women in one vector.
```{r}
#?factor()
```
```{r}
factor(c(1, 1, 2, 1, 1, 2), c(1, 2), c('Men', 'Women'))
```
```{r}
levels(factor(c(1, 1, 2, 1, 1, 2), c(1, 2), c('Men', 'Women')))
```

## Lists

List is a special type of vector, which can contain elements of different classes or, actually, anything you want. It is an extremely versatile and important data type. 

To create a list, you can use function **list()**.
```{r}
list(
  c(1, 2, 3)
)
```
```{r}
list(
  NA,
  c(1,2,5),
  TRUE
)
```

List is a recursive data structure, i.e. they can contain lists inside another lists. It becomes crucial, if you want to parse unstructured data.

```{r}
list(
  list(
    1,
    c(1, 2),
    list(
      c(1, 2, 3)
    )
  )
)
```

## Data frames

Data frames are a special implementation of a list, where each element needs to have the same length. They are primarily used to store tabular data. To create a data frame, you need to use **data.frame()** function.

```{r}
data.frame(
  c(1, 2, 3), #length of 3 elements
  c(2, 3, 4), #length of 3 elements
  c('t', 't', 't') # length of 3 elements
  )
```

```{r, eval=F}
data.frame(
  c(1, 2, 3), #length of 3 elements
  c(2, 3, 4), #length of 3 elements
  c('t', 't', 't', 't') # length of 4 elements, the code will throw an error
  )
```

To name columns of a data frame, you should use '='.

```{r}
data.frame(
  column_1 = c(1, 2, 3),
  column_2 = c(2, 3, 4)
)
```

# Creating a variable

To assign variable name to these data types you need to use assignment operator (<-, Alt+-)
```{r}
vector_1 <- c(1, 2, 3)
data_frame_2 <- data.frame(
  column_1 = c(1, 2, 3),
  column_2 = c(2, 3, 4)
)
```


# Additional materials

[How to name variables](http://adv-r.had.co.nz/Style.html#:~:text=Object%20names&text=Variable%20and%20function%20names%20should,this%20is%20not%20easy!).)
To refresh your memory on this part of the course, you could always read this [cheat sheet](https://rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf)
![](https://www.youtube.com/watch?v=vGY5i_J2c-c&feature=youtu.be&t=4m43s)
