---
title: "Control flow in R"
author: "Alexandr Byzov"
date: "07 09 2020"
output: html_document
---

Control flow is your way to specify the order of instructions in your code. There are two tools to determine the control flow in R: choices and loops. In this class, I will mention only the most basic ones. If you want to know more about [control flow in R see](https://adv-r.hadley.nz/control-flow.html)

# Choices

In the broad sense, choices run logical tests and return values based on their outcome (TRUE - one value, FALSE - another one). In R, you need to specify TRUE conditions and actions. FALSE actions are optional.

Logical operators:

1. == equals
2. != does not equal
3. >,<,>=,<= greater than, less than etc.
4. %in% identifies if an element belongs to a vector or data frame

```{r}
vector_1 <- c(1, 2, 3, 4, 5)
if (length(vector_1) == 5) {
  print("The length of this vector equals 5")
} else {
  print("The length of this vector does not equal 5")
}
```

```{r}
vector_2 <- c(1, 2, 3, 4, 5, 6)
if (length(vector_2) == 5) {
  print("The length of this vector equals 5")
} else {
  print("The length of this vector does not equal 5")
}
```

# Loops

Loops are used to repeat some tasks. For each item in vector, perform_action is called once, updating the item's value each time.

```{r}
vector_2 <- c(1, 2, 3, 4, 5, 6)
for(i in seq_along(vector_2)){
  print(i)
}
```

If you want to store values from loops, you need to create an empty atomic vector, list, matrix, or data frame. Then, you can use indices to store these values.

```{r}
vector_1 <- c(1, 2, 3)
vector_with_results <- c()
for(i in seq_along(vector_1)){
  vector_with_results[i] <- vector_1[i] + 2
}
print(vector_with_results)
```
