---
title: "R packages"
author: "Александр Бызов"
date: "20 09 2020"
output: html_document
---

# Packages in R

R packages are collections of functions that are developed to enhance the basic capabilities of R greatly. To install a package you need to use *install.packages('Package name in quotation marks')* function.
```{r}
#install.packages('tidyverse') # Tidyverse is a collection of packages to work with data
```
To install several packages, you may pass a character vector with packages' names to *install.packages()* function.
```{r}
#install.packages(c('tidyverse', 'jsonlite')) # jsonlite is a package to parse web data in json format
```
When you have installed needed packages, you may use its functions in two ways:
1. Write down package_name::function_name to use needed function
```{r}
#jsonlite::read_json()
```
2. Load package in working memory with *library(package_name_without_quotation_marks)* function
```{r}
library(jsonlite)
#read_json()
```

To learn more about packages [see](https://www.datacamp.com/community/tutorials/r-packages-guide)
[Tidyverse, a collection of packages to work with data](https://www.tidyverse.org/)