---
title: "3. Subsetting"
author: "Alexandr Byzov"
date: "07 09 2020"
output: html_document
---

# Subsetting atomic vectors  

There are many ways to subset atomic vectors. [H. Wickham](https://adv-r.hadley.nz/subsetting.html) describes six ways to subset this type of vectors.

But before you dive into various subsetting techniques, create a vector of six elements:
```{r}
vector_to_subset <- c(5, 6, 7, 8, 9, 10)
```
1. A subset [] with positive integers will return an element or elements by their positions
```{r}
vector_to_subset[1] # NB! R starts counting from 1, not from 0 as Python does
```
```{r}
vector_to_subset[c(1, 3, 5)]
```

2. A subset with negative integers will exclude an element or elements by their positions
```{r}
vector_to_subset[-1]
```
```{r}
vector_to_subset[c(-1, -3, -5)] # You can not mix positive and negative integers in the subset

```

3. Subset with logical vectors will select element or elements where the corresponding logical value is TRUE

```{r}
vector_to_subset[c(T,T,T,F,F,F)]

```
4. Subset with character vectors will return elements with matching names if the vector is named
```{r}
vector_to_subset_2 <- c(a = 1, b = 2)
vector_to_subset_2['a']
```

# Subsetting matrices

Matrix is a two dimensional object with rows and columns, so to subset elements from this data structure, you need to specify both rows and columns.
```{r}
m <- matrix(c(1, 2, 3, 4, 5, 6), ncol = 2)
```

1. A subset of second column and first two rows
```{r}
#matrix[rows, columns]
m[c(1, 2), 2]
```
2. Exclusion of first column
```{r}
#m[empty_to_subset_all_rows, columns]
m[,-1]
```
3. A subset of first and third row with logical statement
```{r}
m[c(T,F,T),]
```
4. A subset with names
```{r}
colnames(m) <- c('first', 'second')
m[,'second']
```

# Subsetting lists

To subset an element of a list, you could use these operators: [], [[]], and $. 

1. [] always returns smaller lists and works under the same principles as [] in vectors 
```{r}
l <- list(c(1, 2, 3), c(1,2,3), c(T,T), F)
```
```{r}
l[1]
```
```{r}
typeof(l[1])
```
```{r}
l[c(1,4)]
```

```{r}
l[-1]
```

2. [[]] and $ are used to extract a single element of a list. It always returns this element

See the difference
```{r}
l[1]
typeof(l[1])
```
```{r}
l[[1]]
typeof(l[[1]])
```
$ is used with named lists
```{r}
named_list <- list(
  element_1 = c(1, 2, 3),
  element_2 = c(2, 3, 4),
  element_3 = T,
  element_12 = c(T,F),
  element_432 = 'test'
)
```
```{r}
named_list$element_1
```

Beware that $ uses partial matching. It means that even if you don't write full name of a vector, R will still try to find it. It is a bad practice, as it tremendously diminishes predictability of a code.
```{r}
named_list$element_4
```

# Subsetting data frames

```{r}
df <- data.frame(
  a = c(1, 2),
  b = c(2, 3), 
  c = c(3, 4)
)
```

Data frames have the characteristics of matrices and lists:

1. If you subset with a single vector inside square brackets [1], data frames are subsetted the same way as atomic vectors
```{r}
df[1]
```
```{r}
df[1:2]
```
```{r}
df[-1]
```
```{r}
df['b']
```

2. If you subset with two vectors inside square brackets[1, 2], data frames are subsetted the same way as matrices
```{r}
df[1,]
```

```{r}
df[1,2]
```

# Subsetting and assignments

All subsetting operators can be combined with assignment to modify selected values of an input vector: this is called subassignment. The basic form is x[i] <- value.
```{r}
vector_1 <- c(1, 2, 3, 4)
print(vector_1)
vector_1[2] <- 5
print(vector_1)
```

